Chemical potential of liquids and mixtures via Adaptive Resolution Simulation

### 2014-07-03 writeDown/pub.jcp/chem-pot-jcp.tex
fix bugs, final version

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}            
\newcommand{\bluec}[1]{{\color{blue} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{ex}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}
\newcommand{\typea}{{\textrm{A}}}
\newcommand{\typeb}{{\textrm{B}}}
\newcommand{\concenttba}{x_{\textrm{TBA}}}

@@

The first order, often mentioned in this work corresponds to the molecular number density $\rho({\vect r})$.}
Moreover we have shown that, because of the necessary conditions, the accuracy in the atomistic region is independent of the accuracy of the coarse-grained model, thus, in the coarse-grained region, one can use a generic liquid of spheres whose only requirement is that it has the same molecular density of the reference system. In the simulation set up, ${\vect F}_{th}$ is calculated via an iterative procedure \recheck{using} the molecular number density in $\Delta$. The iterative scheme consists of calculating \recheck{${\vect F}^{k+1}_{th}({\vect r}) = {\vect F}^{k}_{th}({\vect r}) + \frac{1}{\rho^2_{0}\kappa_{T}}\nabla \rho^{k}({\vect r})$} ($\kappa_{T}$ the isothermal compressibility), and the thermodynamic force is considered converged when the target density $\rho_{0}$ is reached in $\Delta$. As a result, ${\vect F}_{th}({\vect r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force.

@@

with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in \recheck{the original} AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the {Appendix~\ref{app:tmp2}})}. Moreover we have proven numerically, for the case of liquid water, that

@@

\section{Analytic Derivation of  $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We consider a potential coupling between the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where  $\alpha$ and $\beta$ denotes the atom indexes of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$ of molecule $i$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the transition region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume the system to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the transition region). The thermodynamic force for GC-AdResS ($\vect F_\thf$) and for the auxiliary Hamiltonian AdResS ($\vect F^\hadress_\thf$),
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
As shown in Refs.\cite{prl12, prx}, $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $ \omega_\thf$ denotes the integral of the thermodynamic force $\vect F_\thf$. \\

@@

symmetric w.r.t molecule $i$ and $j$, therefore, the Newton's

@@

  &+
  \int_\Delta d \vect r\, \rho_0 S\cdot \Delta L\cdot
  \big[\vect F^\hadress_\thf(\vect r)

@@

The last term is originated by the work done of the ideal piston. This latter is composed by

@@

their total work is zero (as long as the transition region move infinitesimally along $x$).

@@

where $w_\thf^\hadress$ is the integral of the 

@@

In the thermodynamic limit, the equilibrium volume of the AT region maximize the

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
In the auxiliary Hamiltonian, we do not have the term $w_\ext$ in the above
formula {(being the term $w_\ext$ in GC-AdResS, generated by the non-conservative effect of the force interpolation).}
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

@@

\recheck{In order to show the numerical efficiency of our approach, we compare the time taken to do a full GC-AdResS simulation and the time for a thermodynamic integration calculation for
different systems with varied concentration of TBA in water. The total time required for an GC-AdResS simulation consists of 
the time taken to obtain a converged thermodynamic force and the time taken to obtain the coarse-grained chemical potential. 
The time taken to complete TI procedure at each value of $\lambda$ is summed up to obtain the total time. In this work, the TI 
is done in two stages, first the van der Waals  interactions are coupled followed by the electrostatic coupling. At each stage, 21 
equally distributed values of $\lambda$ are used,
therefore, in total 42 simulations were performed to calculate each TI chemical potential value.
% However, in literature, a single stage procedure which consists of coupling 
%  van der Waals and electrostatic interactions together is followed. In order to have a fair comparison, we have compared GC-AdResS time
% to the single-stage TI procedure. Also, GC-AdResS calculations have been done in a more optimized way, by using a spherical atomistic region and 
% a transition region of $1.4 nm$ each. 
In an AdResS simulation, the initial guess of the thermodynamic force largely determines the time for convergence. 
We started with a randomly chosen initial guess
($-42$~kJ/mol, we picked a small value because the TBA molecule is hydrophilic)
for system with the highest mole fraction of TBA. For all the other systems, 
we used the converged thermodynamic force obtained from the first system as an initial guess. The convergence was much 
faster in all the other cases using this approach. Table~\ref{table3} shows the number of iterations required for 
the thermodynamic force convergence in GC-AdResS and total time required for GC-AdResS and TI calculation.
% The total time for the all the calculations is 66 hours in GC-AdResS and 72 hours in TI.
The advantage of GC-AdResS over TI
is that we get two values of excess chemical potential for both solute and solvent in a single calculation, while 
in TI, the whole process has to be repeated to get the excess chemical potential of other component. For very dilute systems ($\concenttba=0.001$), 
however, one has to take a very large systems in GC-AdResS (see the Appendix~\ref{app:tmp1} for system size). It takes a large amount of time for the thermodynamic force to reach convergence, and hence TI is always a better option at such low concentrations with a much 
smaller system size at the same mole fraction.

@@

\section{Current computational convenience: a critical appraisal}
The natural question arising from the discussion above is whether or not GC-AdResS is a more convenient technical tool for calculating $\mu^{ex}$ compared to TI.
Currently the answer is \recheck{neither negative nor positive}, although the current work is the first step towards a potentially positive answer for the future. In fact, the fastest version of AdResS is implemented in the GROMACS code \cite{gromacs}; using the Gromacs version 4.5.1 a speedup of a factor four with respected to full atomistic simulations has been reported for aqueous mixtures~\cite{debashish1,nico-debashish}. In this case GC-AdResS was more convenient than TI because in one simulation one could obtain the chemical potential of each liquid component and at the same time calculate structural properties (e.g. radial distribution functions). However in the successive version of GROMACS 4.6.1 the performance of atomistic simulations (above all of SPC/E water) has been highly improved while the corresponding implementation of AdResS is not optimized yet. At the current state, AdResS can only assure a speed up factor between \recheck{2 and 3} for large systems (30000 molecules) compared to full atomistic simulations \recheck{(except for pure SPC/E water systems)}. As a consequence for the calculation of $\mu^{ex}$, TI is in general computationally less demanding than AdResS . Another point that must be considered (in perspective) for a fair comparison between TI and GC-AdResS, is the following: even if AdResS is optimized, in any code, TI has the advantage that one can use one single molecule in the simulation box to mimic the minor component of a mixture. In our case, instead, we must treat, technically speaking, a true mixture with a certain number of molecules of the minor component immersed in the liquid of the major component. Thus, at low concentrations, GC-AdResS simulations require larger systems than those required by TI, moreover, because of the low density of the minor component, the convergence of the corresponding thermodynamic force requires long simulations. 
Thus, for very dilute systems, if one is interested only in the chemical potential, TI shall be preferred to GC-AdResS, however if the interest goes beyond the calculation of the chemical potential, (e.g. radial distribution functions) then (optimized) GC-AdResS would still be more convenient.
When the concentration becomes higher, GC-AdResS may become preferable for both tasks: general properties of the mixture and chemical potential, not only because in this case one requires larger systems, but also because the convergence of the thermodynamic force of the minor component is much faster. Moreover, we would have the flexibility of calculating the chemical potential
of both components in one simulation run, whereas in TI, one needs to run two separate simulations in order to get the chemical potential of both components.
\recheck{The results reported in the previous section about the current efficiency of GC-AdResS are rather encouraging, however currently there is not a clear convenience in using GC-AdResS instead of TI for calculating $\mu^{ex}$; 
in any case the technical aspects of code optimization must be reported and we must make clear that the aim of this work is to show that  the automatic calculation of $\mu^{ex}$, independently from the simulation code in which is implemented and its computational cost, is a ``conceptual'' feature of GC-AdResS}.

@@

\section*{Acknowledgments}
This work was supported by the Deutsche Forschungsgemeinschaft (DFG) with the Heisenberg grant provided to L.D.S (grant code DE 1140/5-1) and with its associate DFG grants for A.G.(grant code DE 1140/7-1) and for H.W. (grant code DE 1140/4-2).H.W.~and C.S.~thank the financial support by DFG research center MATHEON. \recheck{We thank Christoph Junghans and Debashish Mukherji for contributing to the information reported in section IV about the speed up factor}.

@@

In all the AdResS simulations, the resolution changes only on $x$ direction.
For each system, 30 iterations were performed to obtain a 
converged thermodynamic force and a flat density profile. Each iteration consisted 
of 200 ps of equilibration which was followed by 200 ps of data collection. 
The simulations were performed at NVT conditions where the temperature was kept constant 
at 298 K. Simulations of liquid methane and ethane were performed at
111.66 K and 184.52 K respectively. \recheck{As it was discussed in~\cite{prx}, there is no requirement 
of a coarse-grained model that resembles the structural and thermodynamic properties of a full atomistic
model. It was shown numerically that the proper exchange of energy and molecules was independent from the 
molecular model used in the coarse-grained region, showing the convenience of GC-AdResS. In this work, a generic WCA potential was used in the 
coarse-grained region. The interaction potential between the coarse-grained particles is given by 
\begin{equation}
  U(r) = 4 \epsilon \left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^{6}\right] + \epsilon,
  \quad
  r \leq 2^{1/6}\sigma. 
\end{equation}
The parameters $\sigma$ and $\epsilon$ were chosen such that the radial distribution functions of particles 
reproduce a liquid structure. For water molecule, the parameters used in this study are $\epsilon=0.65$~kJ/mol
and $\sigma=0.30$~nm. Table~\ref{cg} shows the WCA parameters for other molecules used in this work.
For interactions between solute and solvent, $\sigma$ values were obtained by 
averaging over the individual parameters.
The solute-solvent $\epsilon$ is the same as the solute-solute  $\epsilon$.
\\
}

@@

To obtain the chemical potential of coarse-grained component, 
insertion particle method was used, where a trajectory of 8~ns was obtained and the coordinates were written
after every 0.4~ps. The insertions of the molecule were performed 4,000,000 times in each 
frame at random locations and with random orientations of the molecule.  
\recheck{The excess chemical potential value was calculated by averaging over the last ten iterations after 
the thermodynamic force has converged and the statistical uncertainty is determined by the standard deviation in the data. \\}

>>>>

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{black} #1}}
\newcommand{\redc}[1]{{\color{black} #1}}            
\newcommand{\bluec}[1]{{\color{black} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{ex}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}
\newcommand{\typea}{{\textrm{A}}}
\newcommand{\typeb}{{\textrm{B}}}
\newcommand{\concenttba}{x_{\textrm{TBA}}}

@@

The first order, often mentioned in this work, corresponds to the molecular number density $\rho({\vect r})$.}
Moreover we have shown that, because of the necessary conditions, the accuracy in the atomistic region is independent of the accuracy of the coarse-grained model, thus, in the coarse-grained region, one can use a generic liquid of spheres whose only requirement is that it has the same molecular density of the reference system. In the simulation set up, ${\vect F}_{th}$ is calculated via an iterative procedure {using} the molecular number density in $\Delta$. The iterative scheme consists of calculating {${\vect F}^{k+1}_{th}({\vect r}) = {\vect F}^{k}_{th}({\vect r}) + \frac{1}{\rho^2_{0}\kappa_{T}}\nabla \rho^{k}({\vect r})$} ($\kappa_{T}$ the isothermal compressibility), and the thermodynamic force is considered converged when the target density $\rho_{0}$ is reached in $\Delta$. As a result, ${\vect F}_{th}({\vect r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force.

@@

with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). {While the determination of $\omega_{\dof}$ is not required for our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\ext}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in {the original} AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. {In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the {Appendix~\ref{app:tmp2}})}. Moreover we have proven numerically, for the case of liquid water, that

@@

\section{Analytic Derivation of  $\omega_{\ext}$}\label{sec:tmp4}
{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We consider a potential coupling between the atomistic and coarse-grained resolution, that is the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where  $\alpha$ and $\beta$ denotes the atom indices of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$ of molecule $i$.
{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
{We refer to the AdResS simulation using potential scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the transition region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume the system to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the transition region). The thermodynamic force for GC-AdResS ($\vect F_\thf$) and for the auxiliary Hamiltonian AdResS ($\vect F^\hadress_\thf$),
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
As shown in Refs.\cite{prl12, prx}, $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $ \omega_\thf$ denotes the work of the thermodynamic force $\vect F_\thf$. \\

@@

symmetric w.r.t molecules $i$ and $j$, therefore, the Newton's

@@

  &-
  \recheck{\int_{\Delta_{x}} d x\, \rho_0 S\cdot \Delta L\cdot
  \big[\vect F^\hadress_\thf(x)

@@

The last term is originated by the work done by the ideal piston. This term is composed by

@@

their total work is zero (as long as the transition region moves infinitesimally along $x$).

@@

where $\omega_\thf^\hadress$ is the work of the 

@@

In the thermodynamic limit, the equilibrium volume of the AT region maximizes the

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
In the auxiliary Hamiltonian, we do not have the term $\omega_\ext$ in the above
formula {(being the term $\omega_\ext$ in GC-AdResS, generated by the non-conservative effect of the force interpolation).}
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

@@

{In order to show the numerical efficiency of our approach, we compare the time taken to do a full GC-AdResS simulation and the time for a thermodynamic integration calculation for
different systems with varied concentration of TBA in water. The total time required for an GC-AdResS simulation consists of 
the time taken to obtain a converged thermodynamic force and the time taken to obtain the coarse-grained chemical potential. 
The time taken to complete TI procedure at each value of $\lambda$ is summed up to obtain the total time. In this work, the TI 
is done in two stages, first the van der Waals  interactions are coupled followed by the electrostatic coupling. At each stage, 21 
equally distributed values of $\lambda$ are used,
therefore, in total 42 simulations were performed to calculate each TI chemical potential value.
% However, in literature, a single stage procedure which consists of coupling 
%  van der Waals and electrostatic interactions together is followed. In order to have a fair comparison, we have compared GC-AdResS time
% to the single-stage TI procedure. Also, GC-AdResS calculations have been done in a more optimized way, by using a spherical atomistic region and 
% a transition region of $1.4 nm$ each. 
In an AdResS simulation, the initial guess of the thermodynamic force largely determines the time for convergence. 
We started with a randomly chosen initial guess
($-42$~kJ/mol, we picked a small value because the TBA molecule is hydrophilic)
for system with the highest mole fraction of TBA. For all the other systems, 
we used the converged thermodynamic force obtained from the first system as an initial guess. The convergence was much 
faster in all the other cases using this approach. Table~\ref{table3} shows the number of iterations required for 
the thermodynamic force convergence in GC-AdResS and total time required for GC-AdResS and TI calculation.
% The total time for the all the calculations is 66 hours in GC-AdResS and 72 hours in TI.
The advantage of GC-AdResS over TI
is that we get two values of excess chemical potential for both solute and solvent in a single calculation, while 
in TI, the whole process has to be repeated to get the excess chemical potential of the other component. For very dilute systems ($\concenttba=0.001$), 
however, one has to take a very large systems in GC-AdResS (see the Appendix~\ref{app:tmp1} for system size). It takes a large amount of time for the thermodynamic force to reach convergence, and hence TI is always a better option at such low concentrations with a much 
smaller system size at the same mole fraction.

@@

\section{Current computational convenience: a critical appraisal}
The natural question arising from the discussion above is whether or not GC-AdResS is a more convenient technical tool for calculating $\mu^{ex}$ compared to TI.
Currently the answer is {neither negative nor positive}, although the current work is the first step towards a potentially positive answer for the future. In fact, the fastest version of AdResS is implemented in the GROMACS code \cite{gromacs}; using the Gromacs version 4.5.1 a speedup of a factor four with respected to full atomistic simulations has been reported for aqueous mixtures~\cite{debashish1,nico-debashish}. In this case GC-AdResS was more convenient than TI because in one simulation one could obtain the chemical potential of each liquid component and at the same time calculate structural properties (e.g. radial distribution functions). However in the successive version of GROMACS 4.6.1 the performance of atomistic simulations (above all of SPC/E water) has been highly improved while the corresponding implementation of AdResS is not optimized yet. At the current state, AdResS can only assure a speed up factor between {2 and 3} for large systems (30000 molecules) compared to full atomistic simulations {(except for pure SPC/E water systems)}. As a consequence for the calculation of $\mu^{ex}$, TI is in general computationally less demanding than AdResS . Another point that must be considered (in perspective) for a fair comparison between TI and GC-AdResS, is the following: even if AdResS is optimized, in any code, TI has the advantage that one can use one single molecule in the simulation box to mimic the minor component of a mixture. In our case, instead, we must treat, technically speaking, a true mixture with a certain number of molecules of the minor component immersed in the liquid of the major component. Thus, at low concentrations, GC-AdResS simulations require larger systems than those required by TI, moreover, because of the low density of the minor component, the convergence of the corresponding thermodynamic force requires long simulations. 
Thus, for very dilute systems, if one is interested only in the chemical potential, TI shall be preferred to GC-AdResS, however if the interest goes beyond the calculation of the chemical potential, (e.g. radial distribution functions) then (optimized) GC-AdResS would still be more convenient.
When the concentration becomes higher, GC-AdResS may become preferable for both tasks: general properties of the mixture and chemical potential, not only because in this case one requires larger systems, but also because the convergence of the thermodynamic force of the minor component is much faster. Moreover, we would have the flexibility of calculating the chemical potential
of both components in one simulation run, whereas in TI, one needs to run two separate simulations in order to get the chemical potential of both components.
{The results reported in the previous section about the current efficiency of GC-AdResS are rather encouraging, however currently there is not a clear convenience in using GC-AdResS instead of TI for calculating $\mu^{ex}$; 
in any case the technical aspects of code optimization must be reported and we must make clear that the aim of this work is to show that  the automatic calculation of $\mu^{ex}$, independently from the simulation code in which is implemented and its computational cost, is a ``conceptual'' feature of GC-AdResS}.

@@

\section*{Acknowledgments}
This work was supported by the Deutsche Forschungsgemeinschaft (DFG) with the Heisenberg grant provided to L.D.S (grant code DE 1140/5-1) and with its associate DFG grants for A.G.(grant code DE 1140/7-1) and for H.W. (grant code DE 1140/4-2).H.W.~and C.S.~thank the financial support by DFG research center MATHEON. {We thank Christoph Junghans and Debashish Mukherji for contributing to the information reported in section IV about the speed up factor}.

@@

In all the AdResS simulations, the resolution changes only along the $x$ direction.
For each system, 30 iterations were performed to obtain a 
converged thermodynamic force and a flat density profile. Each iteration consisted 
of 200 ps of equilibration which was followed by 200 ps of data collection. 
The simulations were performed at NVT conditions where the temperature was kept constant 
at 298 K. Simulations of liquid methane and ethane were performed at
111.66 K and 184.52 K respectively. {As it was discussed in~\cite{prx}, there is no requirement 
of a coarse-grained model that resembles the structural and thermodynamic properties of a full atomistic
model. It was shown numerically that the proper exchange of energy and molecules was independent from the 
molecular model used in the coarse-grained region, showing the convenience of GC-AdResS. In this work, a generic WCA potential was used in the 
coarse-grained region. The interaction potential between the coarse-grained particles is given by 
\begin{equation}
  U(r) = 4 \epsilon \left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^{6}\right] + \epsilon,
  \quad
  r \leq 2^{1/6}\sigma. 
\end{equation}
The parameters $\sigma$ and $\epsilon$ were chosen such that the radial distribution functions of particles 
reproduce a liquid structure. For water molecule, the parameters used in this study are $\epsilon=0.65$~kJ/mol
and $\sigma=0.30$~nm. Table~\ref{cg} shows the WCA parameters for other molecules used in this work.
For interactions between solute and solvent, $\sigma$ values were obtained by 
averaging over the individual parameters.
The solute-solvent $\epsilon$ is the same as the solute-solute  $\epsilon$.
\\
}

@@

To obtain the chemical potential of coarse-grained component, 
insertion particle method was used, where a trajectory of 8~ns was obtained and the coordinates were written
after every 0.4~ps. The insertions of the molecule were performed 4,000,000 times in each 
frame at random locations and with random orientations of the molecule.  
{The excess chemical potential value was calculated by averaging over the last ten iterations after 
the thermodynamic force has converged and the statistical uncertainty is determined by the standard deviation in the data. \\}
